A Framework for Classification of Social Machines

### 2013-02-14 latex/sig-alternate.tex
Added more about framework, and added motivation figure

While determining a set of social machines, one clear distinction was between social machine frameworks, such as MediaWiki and Ushahidi, which enable
social machines to be created, and instantiations of those frameworks into social machines that have been created for a specific purpose, such as Wikipedia
and Ushahidi Haiti. However, through further investigation, we discovered that the distinction between framework and instantiations was not the only
differentiator. Specifically, that some social machines could also have sub-machines or communities within them, that are working to solve different problems,
within the context of the larger machine. For example, Wikipedia is an instantiation of the MediaWiki framework, but there are a large number of
communities within Wikipedia that work on domain-specific areas, such as those contributing in areas of science and technology\footnote{Visualisation of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}. However, we must also note that wikipedia is primarily split into communities of language, with the English, German and French wikipedias having the largest number of articles. It is therefore possible to analyse wikipedia into a hierarchy, from MediaWiki, to the languages of Wikipedia, down to the domain-specific communities. We explored this idea further, to include other social machines, and links where social machines utilise other social machines in order to function. For example, the social machine used by the Obama '12 US Presidential campaign~\cite{obamakieron}, which relied on social machines such as Twitter and Facebook. A resulting polyarchy is shown in Figure \ref{polyarchy}, which classifies levels as Infrastructure, Frameworks, Services and Causes/Groups.

@@

@@

>>>>

While determining a set of social machines, one clear distinction was between social machine frameworks, such as MediaWiki and Ushahidi, which enable
social machines to be created, and instantiations of those frameworks into social machines that have been created for a specific purpose, such as Wikipedia
and Ushahidi Haiti. However, through further investigation, we discovered that the distinction between framework and instantiations was not the only
differentiator. Specifically, that some social machines could also have sub-machines or communities within them, that are working to solve different problems,
within the context of the larger machine. For example, Wikipedia is an instantiation of the MediaWiki framework, but there are a large number of
communities within Wikipedia that work on domain-specific areas, such as those contributing in areas of science and technology\footnote{Visualisation of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}. However, we must also note that wikipedia is primarily split into communities of language, with the English, German and French wikipedias having the largest number of articles. It is therefore possible to analyse wikipedia into a hierarchy, from MediaWiki, to the languages of Wikipedia, down to the domain-specific communities. We explored this idea further, to include other social machines, and links where social machines utilise other social machines in order to function. For example, the social machine used by the Obama '12 US Presidential campaign~\cite{obamakieron}, which relied on social machines such as Twitter and Facebook.

Looking at the set of social machines in a polyarchy leads to a broad/specific relationship emerging that lets us talk about behaviours at various levels of granularity.  We propose looking at nested machines with ``The Web'' as one of several potential roots, with the next level down consisting of sub-platforms (such as Facebook, Twitter, MediaWiki, Ushahid platforms) that spawn more specific social machines. A resulting polyarchy is shown in Figure \ref{polyarchy}, which classifies levels as Infrastructure, Frameworks, Services and Causes/Groups.

@@

This approach enables us to start with a more detailed analysis of certain levels over others;
and seeing what similarities flow up and down the polyarchy. For example, what do
specific instances of Ushahidi/Zooniverse/MediaWiki have in common with other instances, and
how to they differ dynamically?

\subsection{Multi-faceted motivation, and who-gets-what benefit}

We identified a hierarchy of constructs pertaining to the motivation people have for using social
machines with top level constructs of:

\begin{enumerate}
\item Hedonistic
\item Financial
\item To be informed
\item To help others
\end{enumerate}

The sub constructs are illustrated in Figure \ref{motivation}, and we relate them to Maslow's hierarchy of needs~\cite{maslow}, (e.g., self-actualisation, security, respect of others), in various ways.

\begin{figure*}[htb]
\begin{center}
\includegraphics[width=18cm]{img/motivation.pdf}
\caption{Motivation hierarchy of participants using Social Machines, related to Maslow's hierarchy of needs.} \label{motivation}
\end{center}
\end{figure*}

For each of these motivations, different roles that participate may have different {\it core motivations}, such as:

\begin{enumerate}
\item Benefit to contributor
\item Benefit to moderator
\item Benefit to system operator/host (e.g, Google, Facebook, Amazon)
\item To affiliates of the host (e.g., Amazon affiliates, YouTube partners)
\item To society as a whole
\item To a contributor's social network
\end{enumerate}

@@

[more here]

\subsection{Metrics of participants and usage}

Finally, we identified several contracts that are relevant for all social machines. First, there are construct that relate to ``analytics'' that can be measured at a point in time, and may change over time. Specifically:

\begin{enumerate}
\item Size
\item Maturity
\item Competitors
\item Diversity of user base
\item Popularity over time
\item Frequency of participation
\end{enumerate}

Next, there are constructs regarding the design affordances of the social machine:

\begin{enumerate}
\item Does it afford anonymity
\item Does it afford quality control or moderation, and in what ways
\item What types of contribution does it afford
\item Openness of resulting data set
\end{enumerate}

In the middle we have a mixture of the result of how the design affordances actually were realised by the social computation, such as how users co-opted / incorporated / interpreted the constraints and:

\begin{enumerate}
\item worked within them
\item worked around them
\item gamed them
\item extended and enhanced them
\item derived new structures and conventions (e.g., hashtags)
\end{enumerate}

The result is what we call `participation' constructs including:

\begin{enumerate}
\item Roles emerging in the social machine  (domain expert, troll/spammer, hub, personal gainer)
\item Quality of contributions
\item Content of contributions
\item Global / cultural reach
\item Task reach --- what do people use it for? 	
\item Active/Passive participation roles 
\end{enumerate}
